Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-12 README.qmd
editing text

---
output: github_document
---

<!-- README.md is generated from README.qmd. Please edit that file -->

@@

<!--
[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15847809.svg)](https://doi.org/10.5281/zenodo.15847809)
-->

@@

> Yi Zhe, Ben Marwick, Li Hao, Shuwen Pei (2025). Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)". Submitted to _Journal of Archaeological Science_

@@

  - [:file\_folder: paper](/analysis/paper): Quarto source document for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. 
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations

@@

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:     
- open the `.Rproj` file in RStudio     
- run `renv::restore()` to ensure you have the packages this analysis depends on     
- finally, open `analysis/paper/article.qmd` and render to produce the `article.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

>>>>

---
format: gfm
default-image-extension: ""
editor: 
  markdown: 
    wrap: 72
---

@@

```{=html}
<!--
[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15847809.svg)](https://doi.org/10.5281/zenodo.15847809)
-->
```

@@

> Ye Zhi, Ben Marwick, Li Hao, Shuwen Pei (2025). Spherical harmonic
> analysis of faceted spheroids identifies shaping strategies and
> standardisation at Qianshangying (North China)". Submitted to *Journal
> of Archaeological Science*

@@

-   [:file_folder: paper](/analysis/paper): Quarto source document for
    manuscript. Includes code to reproduce the figures and tables
    generated by the analysis.
-   [:file_folder: data](/analysis/data): Data used in the analysis.
-   [:file_folder: figures](/analysis/paper/figures): Plots and other
    illustrations

The **py_SPHARM** directory contains:

-   [:file_folder: py_SPHARM](/py_SPHARM): Python source files for
    SPHARM analysis.
    -   **SPHARM_main**:
        -   `batch_operation.py` — batch calculation of spherical
            harmonics for STL files.\
        -   `main.py` — shape analysis using spherical harmonics for a
            single STL file, including visualization of analysis
            results.\
    -   **SPHARM_modules**: supporting modules for main scripts,
        including shape preprocessing, spherical harmonic
        computation/statistics, shape reconstruction, and
        computation/output of sphericity and curvature.

@@

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:\
- open the `.Rproj` file in RStudio\
- run `renv::restore()` to ensure you have the packages this analysis
depends on\
- finally, open `analysis/paper/article.qmd` and render to produce the
`article.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")`
in the R console

### 2025-09-12 analysis/paper/paper.qmd
editing text

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). Among these, 13 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage. They are all faceted spheroids with intensive flake scars on the surface. Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.

@@

::: {#fig-Spheroids-distribution-China}
```{r, echo=FALSE}
#| label: fig-SPHARM-steps
#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis"

@@

a. UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b. Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere
:::
As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020]. Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020]. The resulting coefficients were normalized for scale-invariant comparison across models.

>>>>

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage. They are all faceted spheroids with intensive flake scars on the surface. Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.

@@

```{r, echo=FALSE}
#| label: fig-SPHARM-steps
#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis. a. UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b. Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere"

@@

a. UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b. Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere

As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020]. Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020]. The resulting coefficients were normalized for scale-invariant comparison across models.
